---
title: "Does cost recovery lead to asset growth for electric utilities?"
author: "Edgar Aguilar"
institute: "American University"
date: "ea9339a@american.edu"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [robot, robot-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
---   
```{r setup, include=FALSE}
# Global chunk options 
knitr::opts_chunk$set(
  echo = FALSE,
  prompt = FALSE,
  tidy = FALSE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  dev = "svglite",   
  fig.align = "center"
)

# Load required package(s)
library(readxl)
library(knitr)

# Check if the data file exists; adjust the path if necessary.
data_file <- "slides_utilities_data.xlsx"
if (!file.exists(data_file)) {
  stop("Data file '", data_file, "' not found. Please ensure the file exists in your working directory or adjust the file path.")
}

# Load the dataset.
data_slides <- readxl::read_xlsx(data_file)
```

# Overview
- Utilities play a key role in ensuring reliable, high-quality electricity service.

- Asset growth serves as proxy for improvements in infrastructure, access, and service quality.

- This analysis explores whether cost recovery helps explain asset growth.

- I begin by showing how many utilities recover their costs and how many do not.I then examine asset growth by cost recovery levels and test the relationship using regression.

- Results show cost recovery has limited explanatory power for asset growth — other factors likely matter more.

---
# Description of the dataset
- There are `r length(unique(data_slides$"Utility short name"))` unique utilities and `r length(unique(data_slides$Country))` unique countries in the dataset.
- Most power utilities in the sample are in the lower and upper middle income groups, with fewer in low or high income countries.
- I look at Average Cost Recovery (for period 2012-2022) and Annual Asset Growth Rate (for the same period) across these utilities.

```{r bar-chart-utilities, fig.width=7, fig.height=5, echo=FALSE, fig.bg='transparent'}
# Define the desired order for the income groups
income_levels <- c("Low income", "Lower middle income", "Upper middle income", "High income")

# Create a frequency table for the ordered Income Group
utilities_income_count <- table(factor(data_slides$`Income Group`, levels = income_levels))

# Increase the bottom margin to provide enough space for angled labels and x-axis label
old_par <- par(mar = c(10, 4, 4, 2) + 0.1)

# Create the bar chart WITHOUT the built-in x-axis label (xlab = "")
bar_positions <- barplot(
  utilities_income_count,
  main = "Power Utilities by Income Group",
  xlab = "",               # Leave this blank; we'll add manually using mtext()
  ylab = "Count of Utilities",
  col = "steelblue",
  border = "white",
  ylim = c(0, max(utilities_income_count) * 1.15),
  xaxt = "n"               # Suppress automatic x-axis labels
)

# Manually add the angled x-axis labels
text(
  x = bar_positions, 
  y = par("usr")[3] - 0.05 * diff(par("usr")[3:4]),  
  labels = income_levels, 
  srt = 30,         # Rotate labels by 30 degrees
  adj = c(1, 1),    
  xpd = TRUE,       # Allow drawing in the margin
  cex = 0.9         
)

# Manually add the x-axis label *below* the angled category labels
# 'line' controls how far below the plot area the text is drawn
mtext("Income Group", side = 1, line = 5, cex = 1.0)

# Annotate each bar with its count value
text(
  x = bar_positions,
  y = utilities_income_count,
  labels = utilities_income_count,
  pos = 3,         # Places text above each bar
  cex = 0.8,
  col = "black"
)

# Reset the graphical parameters to their original settings
par(old_par)
```


---
# Cost recovery across utilities
- Only 69 (37%) of 189 utilities recover both operational and debt-related costs.
- On average, utilities in high- and upper-middle-income countries achieve cost recovery, while those in low- and lower-middle-income countries fall short.

.pull-left[
```{r pie-chart, fig.width=7, fig.height=5, echo=FALSE}
# Convert "Average Cost Recovery" to numeric if it isn't already
data_slides$`Average Cost Recovery` <- as.numeric(data_slides$`Average Cost Recovery`)

# Create a grouping variable: utilities above 100% recovery vs. below 100%
data_slides$recovery_group <- ifelse(data_slides$`Average Cost Recovery` > 1, "Above 100%", "Below 100%")

# Count utilities in each group
recovery_counts <- table(data_slides$recovery_group)

# Create a pie chart with custom labels and colors
pie(recovery_counts,
    labels = paste0(names(recovery_counts), " (", recovery_counts, ")"),
    main = "Utilities by Average Cost Recovery",
    col = c("lightblue", "lightgreen"))
```
]
.pull-right[
```{r box-plot-cost, fig.width=7, fig.height=5, echo=FALSE}
# Filter out observations with "Average Cost Recovery" above 3
data_filtered <- subset(data_slides, `Average Cost Recovery` <= 3)

# Convert Average Cost Recovery to percentage by multiplying by 100
data_filtered$`Average Cost Recovery (%)` <- data_filtered$`Average Cost Recovery` * 100

# Define the desired order for the Income Group and set the factor levels
desired_levels <- c("Low income", "Lower middle income", "Upper middle income", "High income")
data_filtered$`Income Group` <- factor(data_filtered$`Income Group`, levels = desired_levels)

# Increase the bottom margin to make room for both the angled category labels and the x-axis label
old_par <- par(mar = c(10, 4, 4, 2) + 0.1, cex.axis = 0.9)

# Create the box plot without drawing the x-axis labels (xaxt="n")
bp <- boxplot(`Average Cost Recovery (%)` ~ `Income Group`, 
              data = data_filtered,
              main = "Average Cost Recovery by Income Group",
              xlab = "",  # Remove x-axis label here; we'll add it manually
              ylab = "Average Cost Recovery (%)",
              col = "lightblue",
              border = "darkblue",
              xaxt = "n")

# Draw the x-axis ticks (without labels)
axis(1, at = 1:length(desired_levels), labels = FALSE)

# Manually add the x-axis category labels at a 30° angle
text(x = 1:length(desired_levels), 
     y = par("usr")[3] - 0.05 * diff(par("usr")[3:4]),  # Position slightly below the axis line
     labels = desired_levels, 
     srt = 30,         # Rotate labels by 30 degrees
     adj = c(1, 1),    # Adjust text alignment (tweak as needed)
     xpd = TRUE,       # Allow labels to be drawn in the margin
     cex = 0.9)        # Set text size

# Add the x-axis title below the angled labels with extra space (increase 'line' to move farther from labels)
mtext("Income Group", side = 1, line = 5, cex = 1)

# Compute group means and medians by Income Group
group_means <- tapply(data_filtered$`Average Cost Recovery (%)`, 
                      data_filtered$`Income Group`, mean, na.rm = TRUE)
group_medians <- tapply(data_filtered$`Average Cost Recovery (%)`, 
                        data_filtered$`Income Group`, median, na.rm = TRUE)

# Overlay markers for mean (red circles) and median (blue triangles)
points(1:length(desired_levels), group_means, pch = 19, col = "red")
points(1:length(desired_levels), group_medians, pch = 17, col = "blue")

# Add a legend to the plot in the top right corner
legend("topright", legend = c("Mean", "Median"), pch = c(19, 17), col = c("red", "blue"))

# Reset graphical parameters to their original settings
par(old_par)
```
]
---
# Cost recovery and utility's asset growth
- Average Cost Recovery is not a significant predictor of Total Assets Growth Rate (p = 0.225); the model explains very little variation (R² = 0.008).
- Other factors beyond cost recovery may better explain asset growth.
- For example, access to external financing, government support, or capital investment programs could drive asset expansion regardless of cost recovery performance.

.pull-left[
```{r regression-table, echo=FALSE}
# Run the regression
model <- lm(`Total Assets Growth Rate` ~ `Average Cost Recovery`, data = data_slides)

# Extract regression coefficients and rename columns
reg_table <- as.data.frame(summary(model)$coefficients)
colnames(reg_table) <- c("Estimate", "Std. Error", "t-value", "p-value")

# Print the table using kableExtra with a chosen font size
library(knitr)
library(kableExtra)
kable(reg_table, caption = "Regression Results: Total Assets Growth Rate on Average Cost Recovery", align = "c") %>%
  kable_styling(full_width = FALSE, font_size = 10)
```
]

.pull-right[
```{r box-plot-asset-growth, fig.width=7, fig.height=5, echo=FALSE}
# Ensure "Average Cost Recovery" is numeric
data_slides$`Average Cost Recovery` <- as.numeric(data_slides$`Average Cost Recovery`)

# Create buckets for Average Cost Recovery
data_slides$ACR_bucket <- cut(data_slides$`Average Cost Recovery`,
                              breaks = c(-Inf, 0.5, 0.75, 1, Inf),
                              labels = c("Below 50%", "Between 50%-75%", "Between 75%-100%", "Above 100%"),
                              right = FALSE)

# Filter out observations where Total Assets Growth Rate exceeds 0.40 (i.e., 40%)
data_filtered <- subset(data_slides, `Total Assets Growth Rate` <= 0.40)

# Convert Total Assets Growth Rate to percentage by multiplying by 100
data_filtered$`Total Assets Growth Rate (%)` <- data_filtered$`Total Assets Growth Rate` * 100

# Set extra bottom margin to give room for both the angled labels and the x-axis title.
# Also, set a common text size for axis labels via cex.axis.
old_par <- par(mar = c(10, 4, 4, 2) + 0.1, cex.axis = 0.7)

# Create the box plot while suppressing both x and y axes.
bp <- boxplot(`Total Assets Growth Rate (%)` ~ ACR_bucket, 
              data = data_filtered,
              main = "Total Assets Growth Rate by Average Cost Recovery Bucket",
              xlab = "",                        # Remove built-in x-axis label.
              ylab = "Total Assets Growth Rate (%)",
              col = "lightblue",
              border = "darkblue",
              xaxt = "n", 
              yaxt = "n")

# Manually add the y-axis with specified text size.
axis(2, cex.axis = 0.7)

# Get the bucket positions and labels for the x-axis.
bucket_positions <- 1:length(levels(data_filtered$ACR_bucket))
bucket_labels <- levels(data_filtered$ACR_bucket)

# Draw the x-axis ticks (without labels).
axis(1, at = bucket_positions, labels = FALSE)

# Manually add the x-axis category labels at a 30° angle with reduced text size.
text(x = bucket_positions, 
     y = par("usr")[3] - 0.05 * diff(par("usr")[3:4]),  # Position slightly below the axis line.
     labels = bucket_labels, 
     srt = 30,         # Rotate labels by 30 degrees.
     adj = c(1, 1),    # Adjust alignment (tweak if needed).
     xpd = TRUE,       # Allow labels to be drawn outside the plotting area.
     cex = 0.7)        # Set text size to match the y-axis labels.

# Add the x-axis title manually with extra space from the category labels.
mtext("Cost Recovery Bucket", side = 1, line = 5, cex = 1)

# Compute means and medians for each ACR bucket.
means <- tapply(data_filtered$`Total Assets Growth Rate (%)`, data_filtered$ACR_bucket, mean, na.rm = TRUE)
medians <- tapply(data_filtered$`Total Assets Growth Rate (%)`, data_filtered$ACR_bucket, median, na.rm = TRUE)

# Overlay markers for mean (red circles) and median (blue triangles) on the box plot.
points(bucket_positions, means, pch = 19, col = "red")
points(bucket_positions, medians, pch = 18, col = "blue")

# Add a legend in the top right corner.
legend("topright", legend = c("Mean", "Median"), col = c("red", "blue"), pch = c(19, 18))

# Reset graphical parameters to their original settings.
par(old_par)
```
]
---
# Thanks!

